Automated Analysis of Muscle X-ray Diffraction \\ Imaging with MCMC

### 2015-05-27 paper.tex
Incorporation of Magda's edits and recommendations

Images generated during experiments share several key features, which
serve as challenges or fiduciary marks during analysis. As seen in
Figure~\ref{fig:imgs}, the brightest part of the image background is
occluded by a circular stop. This physical block is put in place to
prevent damage to the detector from the high photon flux seen at the
center of the X-ray beam. Surrounding the blocked region, the
remainder of the image displays an exponentially decaying background.
Interrupting this largely symmetric background are the diffraction
peaks we locate and model.  Our core data analysis pipeline includes
two steps: image segmentation and image modeling with MCMC processes.

@@

The central circular blocked region is first located and acts as a
relative reference point for subsequent operations. Consistent with
experimental design, it is assumed to contain the center of the
diffraction pattern, although the center of the diffraction pattern is
not assumed to be at the center of the blocked region. The central
blocked area is as dark or darker than the background at image edges
and the edge of the background surrounding it is one of the brightest
regions, so the image is first split into areas with values less than
and areas with values greater than two standard deviations above the
mean. This partitioning yields a binary image where the center blocked
region is surrounded by a halo that is the upper end of the pattern
background and occasional dots where diffraction peaks rise more than
two standard deviations above background. This binary image is
converted to a hierarchical contour set with OpenCV. The blocked
region is taken as the inner-most contour and then modeled as the
smallest circle that can contain it.

\magda{This paragraph is tedious and somewhat confusing but I'm
not sure how best to fix it.}
With the central blocked region identified (shown as a dark magenta
overlay in Figure~\ref{fig:workflow}) we identify local maxima and
mask out those found in regions of the image unlikely to provide
diffraction peaks of interest.  After smoothing through convolution
with a Gaussian kernel maxima are identified by treating each 3 by 3
pixel region as a group to prevent stochastic brightness variation
from identifying maxima separated by single pixel gaps. The resulting
local maxima are then accepted as peaks or rejected based on masking
parameters.  We begin with a complete image and first mask a circular
zone centered on the central blocked region. This prevents the edge
between the blocked region and the background from being interpreted
as ring of peaks. Next we mask regions below the 80th percentile,
eliminating maxima in regions dominated by detector noise. Finally, we
mask regions near the image edge to cull partially cropped peaks. The
resulting unmasked area from which we keep maxima is depicted as light
purple overlay in Figure~\ref{fig:workflow}.

@@

\magda{Before we can say the following, we should say that peaks
come in pairs.}
Next, maxima are clustered into pairs of peaks based on their distance
and angle from the center of the blocked region. Starting with those
maxima nearest the blocked region, a corresponding maxima is sought an
equal distance away from the blocked region and located such that the
angle formed by the two maxima and the center of the blocked region is
180$^\circ$. If no matched maxima exists within a 10\% tolerance, the
maxima is discarded as a spurious peak. With peak pairs now identified
(shown as color matched dots in Figure~\ref{fig:workflow}), the
diffraction center is identified by taking the mean location between
peak pairs and the background is fit and subtracted.

The background is subtracted by first masking arcs encompassing peak
pairs and then fitting a double exponential to a radial profile of the
remaining image. An arc swept 12$^\circ$ out on either side of each
peak pair is sufficient to mask the effect of the diffraction peaks on
the background to be generated (shown as a light gray arc under the
peak pairs in Figure~\ref{fig:workflow}). We next calculate a radial
profile of the image, \magda{centered on the center sounds repetitive}
centered on the pattern center and omitting the regions surrounding
diffraction lines. We fit a double exponential function of the form
$background = a+ b e^{-x c} + d e^{-x e}$ to the radial profile. From
these parameters we generate an estimates background image and
subtract it from the real image, allowing us to extract the $d{10}$
peaks as regions of interest (ROIs) unhindered by an overlaid
diffraction background.

@@

\section{Preliminary Evaluation}

We apply our workflow to a test corpus of 1,220 images generated using
X-ray diffraction during insect flight muscle research at Argonne
National Laboratory's BioCAT Beamline. Sample high-quality and
challenging images are shown in Figures~\ref{fig:imgs}A and
\ref{fig:imgs}B. Our technique segments these images into regions of
interest, identifies the approximate locations of the diffraction
peaks, subtracts background from the regions surrounding the peaks,
and computes the distribution of possible parameters which underly the
shown peaks.  We find that our initial image segmentation step
successfully models the center blocked region of each image in greater
than 99\% of images in our test corpus of 1,220 images. The overall
process allows us to calculate peak-to-peak distances to sub-pixel
accuracy with a confidence interval of 90\%.  After marginalizing over
these other parameters we obtain sub-pixel accuracy with a 95\%
confidence interval for the corpus images on which this technique has
been validated.  MCMC sampling combined with image segmentation allows
us to precisely, accurately, and automatically locate the $d_{10}$
peak center and thus calculate the lattice spacing measured by a
diffraction image to within 0.03 nm.

@@

%%%%%%%%%%%
\section{Challenging and Next Steps}
\label{sec:challenges}

@@

\begin{itemize}[noitemsep]
\item Development of a declarative language to describe processing steps will speed use of this technique and ease reproducibility. 
The key question is to define the types of operations that users should be able to specify and how to specify them. Our goal is
to generalize to a broad set of analysis needs for X-ray images of muscle structures.
\item Packaging of this tool chain into a cloud deployable
containerized image will enable trivial scaling to work with larger
datasets. Furthermore, the ability to access to tool chain directly through a web browser 
with automatic back-end deployment of the analysis pipeline will facilitate adotpion. 
\item Application of these techniques to coming ultra-high temporal-resolution images with far lower signal:noise will strain autosegmentation and peak-fitting techniques. 
\end{itemize}

>>>>

Images generated during experiments share several key features, which
serve as challenges or fiduciary marks during analysis. As seen in
Figure~\ref{fig:imgs}, the brightest part of the image background is
occluded by a circular stop. This physical block is put in place to
prevent damage to the detector from the high photon flux seen at the
center of the X-ray beam. Surrounding the blocked region, the
remainder of the image displays an exponentially decaying background.
Interrupting this background are symmetric pairs of diffraction peaks
we locate and model.  Our core data analysis pipeline includes two
steps: image segmentation and image modeling with MCMC processes.

@@

The dark central circular blocked region is first located and acts as
a relative reference point for subsequent operations. Consistent with
experimental design, it is assumed to contain the center of the
diffraction pattern, although the center of the diffraction pattern is
not assumed to be at the center of the blocked region. The edge of the
background surrounding the block is the brightest region, so the image
is first split between areas with values less than and greater than
two standard deviations above the mean. This partitioning yields a
binary image where the center blocked region is surrounded by a halo
of the upper end of the pattern background and occasional dots where
diffraction peaks rise more than two standard deviations above
background. This binary image is converted to a hierarchical contour
set with OpenCV. The blocked region is taken as the inner-most contour
and then modeled as the smallest enclosing circle, shown as dark
magenta in Figure~\ref{fig:workflow}.

\dave{I cut this down and simplified. It can be removed or further
reworked if desired.}
With the central blocked region located, we identify local maxima in 3
by 3 pixel groups after Gaussian smoothing and mask those found in
regions unlikely to provide peaks of interest.  Resulting maxima are
accepted or rejected as peaks based on masking parameters.  Masked
rejection-regions consist of: 1) a circular zone around the central
blocked region where the blocking generates non-peak local maxima, 2)
areas below the 80th percentile where detector noise dominates, 3)
areas near the image edge where peaks are partially cropped. The
resulting unmasked area from which we keep maxima is depicted as light
purple overlay in Figure~\ref{fig:workflow}.

@@

Next, maxima are clustered into peak pairs based on their distance
and angle from the center of the blocked region. Starting with those
maxima nearest the blocked region, a corresponding maxima is sought an
equal distance away from the blocked region and located such that the
angle formed by the two maxima and the center of the blocked region is
180$^\circ$. If no matched maxima exists within a 10\% tolerance, the
maxima is discarded as a spurious peak. With peak pairs now identified
(shown as color matched dots in Figure~\ref{fig:workflow}), the
diffraction center is identified by taking the mean location between
peak pairs and the background is fit and subtracted.

The background is subtracted by first masking arcs encompassing peak
pairs and then fitting a double exponential to a radial profile of the
remaining image. An arc swept 12$^\circ$ out on either side of each
peak pair masks the effect of the diffraction peaks on the background
(shown as a light gray arc under the peak pairs in
Figure~\ref{fig:workflow}). We calculate a radial profile of the image
around the pattern center, omitting diffraction line regions. We fit a
double exponential function of the form $background = a+ b e^{-x c} +
d e^{-x e}$ to the radial profile. From these parameters we generate
an estimated background image and subtract it from the real image,
allowing us to extract the $d{10}$ peaks as regions of interest (ROIs)
unhindered by an overlaid diffraction background.

@@

%%%%%%%%%%%
\section{Preliminary Evaluation}
\label{sec:eval}

We apply our workflow to a test corpus of 1,220 images generated using
X-ray diffraction during insect flight muscle research at Argonne
National Laboratory's BioCAT Beamline. Sample high-quality and
challenging images are shown in Figures~\ref{fig:imgs}A and
\ref{fig:imgs}B. Our technique segments these images into regions of
interest, identifies the approximate locations of the diffraction
peaks, subtracts background from the regions surrounding the peaks,
and computes the distribution of possible parameters which underly the
shown peaks.  We find that our initial image segmentation step
successfully models the center blocked region of each image in greater
than 99\% of images in our test corpus of 1,220 images. The overall
process allows us to calculate peak-to-peak distances to sub-pixel
accuracy with a confidence interval of 90\%. MCMC sampling combined
with image segmentation allows us to precisely, accurately, and
automatically locate the $d_{10}$ peak center and thus calculate the
lattice spacing measured by a diffraction image to within 0.03 nm.

@@

%%%%%%%%%%%
\section{Challenges and Next Steps}
\label{sec:challenges}

@@

\begin{itemize}[noitemsep]
\item Development of a declarative language to describe processing
    steps will speed use of this technique and ease reproducibility.
    The key question is to define the types of operations that users
    should be able to specify and how to specify them. Our goal is to
    generalize to a broad set of analysis needs for X-ray images of
    muscle structure.
\item Packaging of this tool chain into a cloud deployable
    containerized image will enable trivial scaling to work with
    larger datasets. Furthermore, the ability to access to tool chain
    directly through a web browser with automatic back-end deployment
    of the analysis pipeline will facilitate adoption. 
\item Application of these techniques to coming ultra-high
    temporal-resolution images with far lower signal:noise will strain
    autosegmentation and peak-fitting techniques. 
\end{itemize}
